\Large \bf 6.824 Final Project

### 2014-05-12 writeup/paper.tex
Some paper edits, added some citations

\section{Introduction}
The presented project implements a persistent, fault-tolerant, high-performance key/value store. In order to achieve these goals, the system implements at-most-once semantics, relies upon Paxos for reaching consenus, shards the database across multiple replica groups, persistently stores values to disk, and implements protocols for automatically recovering and updating disk contents upon server reboot.  In order to improve performance, the system features two paxos optimizations: a protocol similar to multi-paxos to avoid the dueling leaders problem, and the ability for leaders to send out prepare messages in advance to reduce the required number of RPCs for paxos agreeement.

@@

\section{Design} \label{sec:design}
The design is based upon the code developed throughout the semester for lab 4.  There are therefore three basic components running on each server: Paxos processes to reach consenus for instances, a shard master to determine which groups are responsible for various shards of the database, and a shard key/value storage process to execute client Put/Get requests.  The Paxos instances are used to create a consistent log of operations for both shard reconfiguration and client operations.

@@

\subsection{Paxos Optimizations}
In order to improve the performance of Paxos, we implemented the Multi-Paxos leader selection optimization. This optimization solved the dueling leaders problem, preventing us from hanging forever in a loop between conflicting leaders, and also allowed us to reduce the number of RPC calls that were necessary to come to agreement, by allowing us to skip the Prepare phase of the Paxos protocol for a leader that was constant between instances. The main issue with the Multi-Paxos leader selection is ensuring that leadership information is propogated through the system and ensuring that we never agree as part of two separate partitions when the Prepare phase is skipped. Unfortunately, we discovered a number of implementation bugs with our default Paxos implementation, which we ended up fixing to end up with a fairly reliable Multi-Paxos optimized system that had a selected leader and skipped unnecessary prepare phases. The results were that the number of required RPCs was reduced by about 33\%, and that there were reduced bandwidth requirements due to the singular leadership. There was an issue that a single machine became the primary proposer, potentially limiting it based on disk and processor speeds, however this doesn't appear to be an issue at the anticipated use cases.

\subsection{Persistent Storage}
In order to tolerate servers restarting and to handle databases that cannot fit in memory, all three components of the system store state and data persistently on disk.  To interface with the disk, a Go wrapper for the high-performance C++ key/value libray LevelDB was utilized.  An outline of what each system component stores on disk is presented below.
\begin{itemize}
\item The Paxos processes write the highest accepted value, the highest accepted proposal number, and the highest acknowledged prepare request to disk for each instance.  This ensures that the protocol will operate correctly even if a server restarts (for example, a decided value can never be changed).  In addition to this state, the Paxos processes persistently store the done responses received from peers and the number of the maximum known instance (see the recovery section below).

@@

\textbf{Single server crashing and restarting, disk in-tact}: The
recovery process contacts a random replica in the same replica group.
The recovering server tells the up-to-date server its current state,
and the up-to-date server responds with any state the recoverer is
missing. This works because the server only discards old decisions if
all participants have marked that sequence number as 'done'.

\textbf{Single server crashing and disk lost}: The recovery process
contacts a random replica in the same replica group, and tells the
up-to-date server that it has no state. The two servers open an
auxillary TCP connection to send over the relevant portions of the
database. This avoids RPC overhead, since the database might be large.
The database has an in-order iterator, in case the TCP connection
fails, the recoverer can re-establish the connection and ask the
sender to pick up where it left off. If the sending server fails, then
the recovering server asks a new replica. This allows for a failure
mode where the original server remians locked permanently. We have
decided that this is preferable to sacrificing consistency, and the
system administartor can resolve the problem by rebooting the hing
server.

\textbf{Complete crash of all servers}: If all servers lose their disk
state, then all will restart and begin like a brand new deployment. If
the disks are not lost, then the recovery process will read the paxos
state written to disk and pick up where it left off. Because we write
to disk before any operation commits, data is not lost.

@@

%% Firstly, the service must be able to handle database content which is too large to fit in volatile memory.  Towards this end, values are written to disk immediately as described above so it does not need to be kept in memory.  More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.  Therefore, a shard transfer occurrs with the following protocol:
%% \begin{itemize}
%% \item The requester sends a request to a peer, including a request number, and the current configuration number.  If doing a shard transfer, the shard number is also included.
%% \item If the receiver is currently at the appropriate configuration, it accepts the request.  It will then begin populating a reply with client response data, a list of seen operations, and the desired key/value data.  While doing this, it will periodically check memory usage; if it exceeds a configurable threshold, it will reply with whatever it currently has and mark the reply as incomplete.
%% \item The requester copies the data from the reply to its disk.  If the reply was marked as incomplete, it sends another request to the same peer with an incremented request number.
%% \item This continues until the reply is marked as complete, at which point the requester launches a goroutine to make sure the sender receives an acknowledgement of the receipt of the final data message.
%% \item If at any point the requester decides that the contacted peer is dead, it launches an acknowledgement goroutine as previously described for that peer (to terminate the transfer in case the peer wakes up or is not dead) and moves on to try a different peer, resetting the request number to 0.
%% \end{itemize}
%% Both the requester and the sender ignore client requests until the transfer is complete - this ensures that the requester will receive a consistent snapshot of the data (and is the reason the requester launches acknowledgement goroutines since otherwise a peer erroneously declared dead would be halted forever).

%% One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.  For example, if a replica group has three servers and a shard transfer is ocurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.  This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.  Since the Paxos instances are stored persistently on disk, the two occupied servers can catch themselves up from the log when they complete the transfer.

@@

\emph{shardkv:} Shardkv specific tests which include deterministically restarting servers to ensure that persistence and recovery operate correctly.  There is also a test which performs the previously written tests but with random reboots of random servers or entire groups.  There are also benchmarks for measuring shard transfer speed.

@@

Adding more groups, however, increases throughput.  For evenly distributed traffic, going from one group to 3 nearly triples the throughput when using multiple keys (thereby leveraging the additional groups).  Figure 2 shows that this trend continues, such that adding groups linearly improves the throughput of the system. The table shows that average throughput for a single client may suffer, probably because the system throughput benchmark creates many clients, while the client throughput test creates a single client; having multiple clients increases the liklihood that a client must wait to see its request served since the replicas may be busy serving another client's request.

@@

One notable bottleneck is disk reading and writing.  The database library wrapper, LeveDB, offers a cache as well as compression and is quite fast.  However, as discussed previously, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.  Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten; this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.  In the future, the system topology could be somewhat altered to allow for more parallelizable recovery and transfer procedures to help minimize the effect of disk latency (for example, a topology inspired by FDS).

Another bottleneck is network capacity and/or number of messages.  As shown in the latency and thoughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.  The paxos optimizations therefore helped to improve performance, and further optimizations could be made to the data transfer protocols to reduce the required number of messages.

Increasing the number of replica groups increases the performance of the system for normal operation, especially under heavy load with evenly distributed key requests. This is especially noticable in the system throughput metric.  This arises since number of groups roughly corresponds to how parallelizable the operations are.

@@

It is also suspected that using UDP would improve throughput, as servers that do not recieve a response already re-transmit their request, making the TCP acknowledgement system redundant.

@@

%{\footnotesize \bibliographystyle{acm}
%\bibliography{paper}}
%\newpage
\end{document}

>>>>

\section{Introduction}
The presented project implements a persistent, fault-tolerant, high-performance key/value store. In order to achieve these goals, the system implements at-most-once semantics, relies upon Paxos for reaching consensus, shards the database across multiple replica groups, persistently stores values to disk, and implements protocols for automatically recovering and updating disk contents upon server reboot.  In order to improve performance, the system features two paxos optimizations: a protocol similar to multi-paxos to avoid the dueling leaders problem, and the ability for leaders to send out prepare messages in advance to reduce the required number of RPCs for paxos agreement.

@@

\section{Design} \label{sec:design}
The design is based upon the code developed throughout the semester for lab 4.  There are therefore three basic components running on each server: Paxos processes to reach consensus for instances, a shard master to determine which groups are responsible for various shards of the database, and a shard key/value storage process to execute client Put/Get requests.  The Paxos instances are used to create a consistent log of operations for both shard reconfiguration and client operations.

@@

\subsection{Paxos Optimizations}
In order to improve the performance of Paxos, the system features the Multi-Paxos leader selection optimization. This optimization solved the dueling leaders problem, preventing infinite loops between conflicting leaders, and also reduced the number of RPC calls necessary to reach agreement by skipping the Prepare phase of the Paxos protocol for a leader that was constant between instances. The main issue with the Multi-Paxos leader selection is ensuring that leadership information is propagated through the system and ensuring that the protocol never agrees as part of two separate partitions when the Prepare phase is skipped. Unfortunately, a number of bugs were discovered with the pre-optimized Paxos implementation, which were fixed before producing a reliable Multi-Paxos optimized system that has a selected leader and skips unnecessary prepare phases.  Once completed, the number of required RPCs was reduced by about 33\%, and bandwidth requirements were reduced due to the singular leadership. There was an issue that a single machine became the primary proposer, potentially limiting the system based on its disk and processor speeds, however this doesn't appear to be an issue at the anticipated use cases.

\subsection{Persistent Storage}
In order to tolerate servers restarting and to handle databases that cannot fit in memory, all three components of the system store state and data persistently on disk.  To interface with the disk, a Go wrapper \cite{levigo} for the high-performance C++ key/value library LevelDB \cite{leveldb} was utilized.  An outline of what each system component stores on disk is presented below.
\begin{itemize}
\item The Paxos processes write the highest accepted value, the highest accepted proposal number, and the highest acknowledged prepare request to disk for each instance.  This ensures that the protocol will operate correctly even if a server restarts \cite{paxos} (for example, a decided value can never be changed).  In addition to this state, the Paxos processes persistently store the done responses received from peers and the number of the maximum known instance (see the recovery section below).

@@

\textbf{Single server crashing and restarting, disk intact}: As described above, each
recovery process contacts a random replica in the same replica group.  The recovering server can then get any missing data by comparing states with the updated server (e.g. shard data, shard master configurations, and Paxos instances - note that Paxos only discards old decisions if all participants have marked that sequence number as 'done').

\textbf{Single server crashing and disk lost}:  This is essentially the same as if the disk was intact, except that much more data will need to be transferred (the same protocol can be used).  Note that when data is transferred, multiple messages may be used for a single shard if it is too large to fit in memory - the sender will periodically check memory usage while constructing the reply and mark replies as incomplete if needed.

% Is the below actually implemented?
%The recovery process contacts a random replica in the same replica group, and tells the up-to-date server that it has no state. The two servers open an auxiliary TCP connection to send over the relevant portions of the database. This avoids RPC overhead, since the database might be large. The database has an in-order iterator, in case the TCP connection fails, the recoverer can re-establish the connection and ask the sender to pick up where it left off. If the sending server fails, then the recovering server asks a new replica. This allows for a failure mode where the original server remains locked permanently. We have decided that this is preferable to sacrificing consistency, and the system administrator can resolve the problem by rebooting the hung server.

\textbf{Complete crash of all servers}: If all servers lose their disk state, then all will restart and begin like a brand new deployment (the above protocols will simply not find any updated data to transfer). If the disks are not lost, then the recovery process will read the various
states written to disk and pick up where it left off.  Because disk writes take place before any operation commits, data is not lost.

It should be noted that during a shard transfer (including recovery), the shard key/value service on both the sender and recoverer ignore client requests in order to ensure that consistent data is sent.  Since their Paxos processes continue running independently, however, progress can still be made even if a majority of servers is involved in a data transfer (the decided operations will be persistently logged and the halted servers can catch up when transfer is complete).

@@

%% Firstly, the service must be able to handle database content which is too large to fit in volatile memory.  Towards this end, values are written to disk immediately as described above so it does not need to be kept in memory.  More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.  Therefore, a shard transfer occurs with the following protocol:
%% \begin{itemize}
%% \item The requester sends a request to a peer, including a request number, and the current configuration number.  If doing a shard transfer, the shard number is also included.
%% \item If the receiver is currently at the appropriate configuration, it accepts the request.  It will then begin populating a reply with client response data, a list of seen operations, and the desired key/value data.  While doing this, it will periodically check memory usage; if it exceeds a configurable threshold, it will reply with whatever it currently has and mark the reply as incomplete.
%% \item The requester copies the data from the reply to its disk.  If the reply was marked as incomplete, it sends another request to the same peer with an incremented request number.
%% \item This continues until the reply is marked as complete, at which point the requester launches a goroutine to make sure the sender receives an acknowledgement of the receipt of the final data message.
%% \item If at any point the requester decides that the contacted peer is dead, it launches an acknowledgement goroutine as previously described for that peer (to terminate the transfer in case the peer wakes up or is not dead) and moves on to try a different peer, resetting the request number to 0.
%% \end{itemize}
%% Both the requester and the sender ignore client requests until the transfer is complete - this ensures that the requester will receive a consistent snapshot of the data (and is the reason the requester launches acknowledgement goroutines since otherwise a peer erroneously declared dead would be halted forever).

%% One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.  For example, if a replica group has three servers and a shard transfer is occurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.  This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.  Since the Paxos instances are stored persistently on disk, the two occupied servers can catch themselves up from the log when they complete the transfer.

@@

\emph{shardkv:} Shardkv specific tests which include deterministically restarting servers to ensure that persistence and recovery operate correctly.  There is also a test which performs the previously written tests but with random reboots of random servers or entire groups.  There are also benchmarks for measuring shard transfer speed and to ensure that the simulated memory limit is never exceeded.

@@

Adding more groups, however, increases throughput.  For evenly distributed traffic, going from one group to 3 nearly triples the throughput when using multiple keys (thereby leveraging the additional groups).  Figure 2 shows that this trend continues, such that adding groups linearly improves the throughput of the system. The table shows that average throughput for a single client may suffer, probably because the system throughput benchmark creates many clients, while the client throughput test creates a single client; having multiple clients increases the likelihood that a client must wait to see its request served since the replicas may be busy serving another client's request.

@@

One notable bottleneck is disk reading and writing.  The database library wrapper, LeveDB, offers a cache as well as compression and is quite fast.  However, as discussed previously, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.  Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten; this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.  It was considered adding a commit point marker (similar to the ideas presented in \cite{harp}), which would be straightforward in the current system by simply returning before the disk write and not advancing the Paxos min until the write completes, but the above measurements indicate that database scanning is much more significant than isolated accesses.  In the future, the improvements can be made to avoid this iteration bottleneck.  For example, the system topology could be altered to allow for more parallelizable recovery with a procedure inspired by FDS \cite{fds}.  More simply, however, the servers can be altered to store a separate database for each shard; the entire database could then be transferred to the receiver using TCP/UDP.  Preliminary tests on AWS showed that a 100MB database can be sent over TCP in less than one second.  In addition, this would avoid the need for either the sender or the receiver to scan through a database, thus eliminating the major bottleneck identified previously.

Another bottleneck is network capacity and/or number of messages.  As shown in the latency and throughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.  The paxos optimizations therefore helped to improve performance, and further optimizations could be made to the data transfer protocols to reduce the required number of messages.

Increasing the number of replica groups increases the performance of the system for normal operation, especially under heavy load with evenly distributed key requests. This is especially noticeable in the system throughput metric.  This arises since number of groups roughly corresponds to how parallelizable the operations are.

@@

It is also suspected that using UDP would improve throughput, as servers that do not receive a response already re-transmit their request, making the TCP acknowledgement system redundant.

@@

%{\footnotesize \bibliographystyle{acm}
%\bibliography{paper}}
%\newpage
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,paper}
\end{document}
